Performance Models for Data Transfers: A Case Study with Molecular Chemistry Kernels

### 2019-04-11 ICPP-19/communications.tex
minor changes

	\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

	%
	% The abstract is a short summary of the work to be presented in the article.
	\begin{abstract}
		With increasing complexity of hardwares, systems with different memory nodes are ubiquitous in high performance computing (HPC). It is paramount to develop strategies to overlap the data transfers between memory nodes with computations in order to exploit the full potential of these systems. In this paper, we consider the problem of deciding the order of data transfers between two memory nodes for a set of independent tasks with the objective to minimize the makespan. We prove that with finite memory capacity, obtaining the optimal order of data transfers is an NP-complete problem. We propose several heuristics for this problem and provide details about their favorable situations. We present analysis of our heuristics on traces, obtained by running 2 molecular chemistry kernels, namely, Hartree–Fock (HF) and Coupled Cluster Single Double (CCSD) on 10 nodes of an HPC system. Our results show that some of our heuristics achieve significant overlap for moderate memory capacities and are very close to the lower bound of makespan.
	\end{abstract}

@@

	With the advent of multicore, and the use of accelerators, it is notoriously cumbersome to exploit the full capability of a machine. Indeed, there are several challenges that come into picture. First, every architecture provides its own efficiency and interface. Therefore, a steep learning curve is required for programmers to take good utilization of all resources. Second, scheduling is a well known NP-Complete optimization problem, and hybrid and distributed resources make this problem harder (we refer ~\cite{webpagescheduling} for a survey on the complexity of scheduling problems and ~\cite{bleuse2015scheduling} for a recent survey in the case of hybrid nodes). Third, due to shared buses and parallel resources, it is challenging to obtain a precise model based on prediction of computation and communication timings. Fourth, then number of architectures has increased drastically in recent years, therefore it is almost impossible to develop hand tuned optimized code for all these architectures. All these observations led to the development of different task based runtime systems. Among several runtimes, we may cite QUARK~\cite{YarKhan:2011:Quark:Manual} and PaRSEC~\cite{parsec} from from ICL, Univ. of Tennessee Knoxville (USA), StarPU~\cite{starpu} form Inria Bordeaux (France), Legion~\cite{legion12} from Stanford Univ. (USA), StarSs~\cite{ompss} from Barcelona Supercomputing Center (Spain), KAAPI~\cite{kaapi} from Inria Grenoble (France). All these runtime systems allow programmers to express their algorithms at the abstract level in the form of direct acyclic graphs (DAG), where vertices represent computations and edges represent dependencies among them. Sometimes some static information is also provided along with the DAG, such as distance to exit (last) node as a priority or affinity of computation towards resources. The runtime is then responsible for managing scheduling of computations and communications, data transfers among different memories, computation-communication overlap, and load balance.

@@

	Certain applications such as dense linear algebra kernels have regular structure. Therefore, it is possible to associate priorities to computations, based on the task graph structure, and to use them at runtime to make the execution efficient. In irregular applications, programmers do not know the precise structure of the task graphs in advance: tasks are added recursively based on certain sentinel constraints. For such applications, the runtime system sees a set of independent ready tasks and schedules them on different processing units. It is extremely important for runtimes to decide the order of data transfers for these scheduled computations so as to maximize the overlap between computations and communications. This is the main topic of this paper. We prove that the order of communications on two memory nodes with the objective of minimizing the makespan is a NP-Complete problem if the memory of the target node is limited. Our proof is inspired from work by~\cite{Papadimitriou:1980:FSL:322203.322213}, which applies a similar technique for 2-machine flowshop problem with bounded capacity. The main difference between both approaches is that they consider that all tasks have the same occupation on the second machine) and that the memory occupation starts when the second machine starts working. On the contrary, our approach is designed for tasks appearing in scientific workloads whose memory requirements are highly irregular and we consider that memory is acquired before starting the data transfer on communication resource. We propose different runtime strategies in order to maximize the overlap of computations and communications. We evaluate our strategies on the context of a cluster of homogeneous nodes. However, our approach is generic and easily adaptable to any system which operates on different memory spaces. Here are the important contributions of this paper:

@@

	The problem of scheduling tasks has been highly studied in the literature and many formulations are known to be NP-Complete~\cite{GareyJohnson}. Many static and dynamic strategies have been proposed and analyzed for scheduling tasks on heterogeneous resources~\cite{heft-Topcuoglu,hipc16multiresource,ipdps16starpu}. There is also a number of studies in the direction of task scheduling with the emphasis of improving locality and minimizing the communication cost~\cite{starpu,heft-Topcuoglu}. Stanisic et. al~\cite{luka-dmdar} proposed a heuristic to schedule tasks on a computing resource where most of its data is available. A similar approach has been adopted by Agullo et. al for the scheduling of sparse linear algebra kernels~\cite{agullo_fmm}. Predari et. al proposed heuristics to partition the task graph across a number of processors such that inter-processor communication can be minimized~\cite{predari:tel-01518956}.

@@

	\noindent\textbf{Problem $DT$} : A set of tasks $ST=\{T_1,
	\cdots, T_n\}$ is scheduled on a processing unit $P$ with
	memory unit $M$ of capacity $C$. Input data for tasks of $ST$
	reside on another memory unit $M'$. $COMM_i$ is the time to
	transfer input data from $M'$ to $M$ for task $i$ and $COMP_i$
	is the execution time of task $i$ on $P$. We assume that these
	tasks do not produce any output data. There can be only one
	communication at a time, and $P$ can only process one task at
	a time. A task uses an amount of memory in memory $M$ from the
	start of its communication to the end of its computation.

@@

	Given a schedule, \scomm($i$) and \scomp($i$) represent the
	start times of task $i$ on communication and computation
	resources. A schedule is feasible if for every time $t$, the
	amount of memory required by tasks such that $\scomm(i) \leq t
	\leq \scomp(i) + COMP_i$ is not more than the memory capacity
	$C$. For simplicity, we assume in all of this paper that the
	amount of memory required by a task is proportional to its
	communication time, and without loss of generality we consider
	that memory size of a task is equal to its communication time.

@@

	Algorithm~\ref{alg:OrderOfExecutionInfinteMemory} presented in Section~\ref{sec:theoreticalProof} achieves an optimal makespan when there is no memory constraint. This optimal value indicates a lower bound on the makespan of the constrained case. We use denote this value with $optimal\_ \allowbreak makespan\_  \allowbreak infinite\_  \allowbreak memory$ ($OMIM$). In the present Section, we propose different heuristics for the limited memory case, and we assess their efficiency with respect to this lower bound in Section~\ref{sec:exp}.

@@

	\subsection{Static Ordering}
	In this class of strategies, we compute the order of processing in advance based on criteria such as communication time, compute time and accelerated tasks. After computing the order, we follow the same order on computation and communication resources and make sure that the memory constraint is respected at each point in the schedule.

@@

	In order to highlight the different behavior of these static strategies, we propose on Table~\ref{tab:staticOrderExample} an example instance, and on Figure~\ref{fig:staticOrderExample} the corresponding schedules for all these heuristics.

@@

	Dynamic strategies are based on the following principle: when the communication resource is idle, a task is chosen based on a ranking criterion which differ depending on the heuristic, among those which fit in memory. For example, if the selection criteria is to choose a highly compute intensive task, then we compute the ratio of compute time and communication time for all tasks, and we select the task which fit in the currently available memory and whose ratio is maximum (ties are broken arbitrarily). We also ensure that the order on both resources is the same, by ordering tasks on the computing resource accordingly.

	\todo[inline]{LED: I added this part ``among those which fit in memory'', is it correct?}
	\todo[inline]{LED: What happens if no task fit in memory?}

@@

	In all strategies, the initial order is obtained by Algorithm~\ref{alg:OrderOfExecutionInfinteMemory}. . We define the following heuristics based on how we select a task from the set of tasks which fit in memory.
	\todo[inline]{LED: What happens if no task fit in memory?}

@@

		\todo[inline]{SKumar: Move to an extended version and more clear about results and implementation}
		Figure~\ref{fig:iterativeLpSolution} shows the performance of different heuristics with MILP based heuristics for various memory capacities of a single trace file.
		%%	\subsection{MILP Based Strategy}
		%%We use GLPK solver v4.65 to solve the MILP formulation described in Section~\ref{subsec:linearprogrammingformulation}. The solver was unable to handle number of variables required to solve MILP at the scale of our interest. Hence, we solve the linear program iteratively for a small subset of size $k=3,4,5$ (solver was unable to handle number of variables required for $k\ge6$ ) and represent the makespan calculated by this heuristic as $lp.k$. At the boundary of two iterations we fixed the event (communication or computation) of an unfinished task who has started before the boundary point and consider other events flexible. We compute various $lp.k$ values for different memory capacities and observe that most of the other heuristics perform better than this heuristic. Figure~\ref{fig:iterativeLpSolution} shows the comparison of different heuristics with MILP based heuristic for various memory capacities of a single trace file.
		\begin{figure}[htb]
			\includegraphics[scale=0.5]{./results/makespan_with_lp.pdf}
			\caption{Comparision of different heuristics with MILP solution based heuristic.\todo[inline]{LED: I understand that this figure will be removed? Otherwise, use aes(fill=) instead of aes(color=) to have color filled bars.}}
			\label{fig:iterativeLpSolution}
		\end{figure}

@@

		Here moderate memory capacity refers to the case where memory is limited but close to the maximum memory requirement of the considered application\todo[inline]{total memory requirement of all tasks?}. We can clearly observe the favorable scenarios for a few heuristics in Figures~\ref{fig:ratio_to_optimal_hf} and ~\ref{fig:ratio_to_optimal_ccsd}. For example, HF compute intensive tasks have small communication timings, hence $SCMR$ heuristic exhibits very good performance in limited memory cases. CCSD has significant percentage of both types of tasks, therefore $OOLCMR$ and $OOSCMR$ heuristics performance is very close to optimal in moderate memory cases. 
		\todo[inline]{LED: Since this is based on experimental results, should this go to a Section 5.5 or 5.6?}

@@

		We consider a machine called Cascade\todo{cite or url?}, available at PNNL, for our experiments. We obtain traces by running two molecular chemistry applications, double precision version of Hartree-Fock (HF) and Coupled Cluster Single Double (CCSD) of NWChem\todo{cite or url} package on 10 nodes of this machine. Each node is composed of 16 Intel Xeon E5-2670 cores. NWChem takes advantages of a Partitioned Global Address Space Programming Model called Global Arrays (GA)\todo{cite or url?} to use shared-memory programming APIs on distributed memory computers. GA dedicates one core of each node to handle other cores, hence we can view a node as being composed of 15 computational cores. We use 150 processes for each application and obtain 150 trace files. We run CCSD with Uracil molecules input and HF with SiOSi molecules (for Uracil molecules, HF has a much smaller workload, each processor executes only around 20 tasks, that is why we chose SiOSi input for HF execution). Each process executes around 300-800 tasks. Our data transfer model is quite simple and we consider that all data transfers between the local memory of each process and the GA memory take the same route. Modeling of different routes of data transfers for the same source-destination pair, bandwidth sharing for different source-destination pairs and network congestion is more challenging and part of our future work. This simple model is enough to provide insight to the application developers (or runtime system) about the ordering of data transfers for the same source-destination pair so as to maximize communication-computation overlap. Our model is easily adaptable to any source-destination pair when there is one fixed route between source and destination (such as between CPU and GPU, one copy engine to transfer data from CPU (resp. GPU) to GPU (resp. CPU) ).
		\todo[inline]{More about tiling}
		Both applications mainly perform two types of computations, tensor transpose and tensor contraction. HF expects to specify a tile size and we set it to 100, so that each core can be used efficiently. CCSD automatically determines tile sizes at different program points based on the input molecules. Hence, HF operates on almost homogeneous tiles while CCSD uses more heterogeneous tiles.

@@

		\begin{figure}[htb]
			%\begin{tabular}{c}
			\includegraphics[scale=0.5]{./results/plots/application_properties.pdf}
			\caption{HF and CCSD tasks characteristics.\todo[inline]{ylab(``Ratio to optimal'') (or to OSIM). Ideally remove underscores from the X axis.}}
			\label{fig:ApplicationProperties}
		\end{figure}

@@

		To get more insights into the considered workloads, we provide in Figure~\ref{fig:ApplicationProperties} information about
		the instances we consider. For each instance, we compute the sum of communication times (\textit{sum comm}) and sum of computation times  (\textit{sum comp}), and normalize it relatively to the $OMIM$ value. Figure~\ref{fig:ApplicationProperties} also shows the maximum of both values, which is a lower bound on the possible makespan of a schedule for this instance, and their sum, which is an upper bound: this represents the makespan of the sequential schedule, obtained with zero overlap between computation and communication). We can see that HF is a communication intensive application and at most 20\% overlap can be expected in the best scenario. On the other hand, in the CCSD workload, communications and computations are almost evenly distributed and a more significant overlap is possible.

@@

		%%	\todo[inline]{Scheduled based on Gilmore-Gomory algorithm and its performance }
		We also consider two other static heuristics for evaluation. The first heuristic is based on an algorithm, proposed by Gilmore and Gomory, to obtain the minimal cost sequence for a set of jobs~\cite{Gilmore-Gomory:1964}. This is a classic algorithm for 2-machine no-wait flow shop. In this algorithm, each job has a start and end state and a cost is associated to change the state. In our context, this cost can be seen as non-overlap time of computation for two adjacent tasks. Here is the main idea of this algorithm. Initially, a partial sequence of jobs is represented by a graph such that their overlap is maximum. Subsequently edges are greedily added to this graph to connect two components while minimizing the total non-overlap cost. When the graph  connected, an edge interchange mechanism is used to determine the sequence of jobs, which ensures that the sequence has minimal cost. More details can be found in the original paper~\cite{Gilmore-Gomory:1964} and our implementation is publicly availabl~\cite{gitworkrepo}. This algorithm does not take memory constraints into account and only provides the sequence of processing. We use this sequence with a memory capacity restriction just like for other static heuristcs, and we call this heuristic \textit{Gilmore-Gomory} ($GG$).

@@

		\subsection{Experimental results}
		We evaluate our scheduling heuristics for several memory capacities. From the obtained traces, we first determine the minimum requirement of the memory capacity $m_c$ to execute all tasks. Then we observe the behavior of all heuristics with memory capacity $m_c$ to $2m_c$, in increments of $0.125m_c$. Our performance metric is the ratio to optimal $r$: if heuristic $H$ has makespan $M_H$ on an instance, and the optimal makespan for the infinite memory case is $OMIM$, then $r(H)=\frac{M_H}{OMIM}$ (lower values are better). This ratio is at least $1$, and values close to $1$ indicate a well-suited  heuristic which achives maximum possible communication-computation.

@@

		As indicated above, HF tasks operate on less heterogeneous tiles, this is also noticeable in Figure~\ref{fig:ratio_to_optimal_hf}. All heuristics depict similar behavior for minimum memory capacity $m_c$ and increasing the memory capacity slightly does not change the performance of all heuristics. As memory capacity is increased further, dynamic variants of heuristics start performing better. For the moderate memory capacities (close to $2m_c$), static order with dynamic correction variants outperform others. $GG$ heuristic does not achieve good performance, because its task sequence is obtained considering no extra memory is available\todo{you mean infinite memory?}, but is then applied in a different scenario where memory is limited. Suprisingly, the $BP$ heuristic which considers only memory constraint obtains good performance for a static heuristic, but is outperformed by more dynamic approaches.

@@

		Figure~\ref{fig:ratio_to_optimal_best_hf} shows the performance comparison of the best variant in each category, in addition to the \textit{order of submission} ($OS$) strategy which orders tasks in the (arbitrary) order in which they are given. Static strategies are expected to perform better when there is not any memory capacity restriction, and indeed this plot shows that static strategies face capacity bottleneck and underperform with limited memory. Dynamic strategies achieve better performance with limited memory capacity, and static order with dynamic correction strategies are suitable for scenarios with moderate memory capacity.\todo[inline]{LED: I do not understand, on Figure~\ref{fig:ratio_to_optimal_best_hf} it seems that \textit{static dynamic best} is always the best option.}

@@

		Figure~\ref{fig:ratio_to_optimal_best_ccsd} shows that best variants of dynamic and static order with dynamic corrections strategies achieve similar performance at minimum memory capacity $m_c$. But as memory capacity increases, heterogeneity allows static order with dynamic corrections based strategies to take advantage of static knowledge to get maximum overlap and dynamic correction to select another task in case of memory capacity limitation. Static strategies also start performing better at the end, which indicates that this application has potential for significant communication-computation overlap and pure dynamic strategies are unable to take this information into account while making scheduling decisions.

@@

		In this paper, we consider the problem of deciding the order of data transfers between two memory nodes such that overlap of communication and computation is maximized. With Exascale computing, applications face bottlenecks due to communications. Hence, it is extremely important to achieve the maximum overlap of computation and communication in order to exploit the full potential of the system. We show that determining the order of data transfers is an NP complete problem. We propose several data transfer heuristics and evaluate them on two molecular chemistry kernels, HF and CCSD. Our results show that some of our heuristics achieve significant overlap and perform very close to the makespan lower bound. We plan to evaluate our strategies on different applications coming from multiple domains. We also plan to study the behavior of our strategies in the context of overlapping CPU-GPU communications with computations. A runtime system aiming at exposing different heuristics to maximize the communication-computation overlap at the developer level and automatically selecting the best one is currently underway.

@@

			We are grateful to Ajay Panyala for his help to run NWChem on Cascade machine. This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research .......
			%% 			Identification of funding sources and other support, and thanks to individuals and groups that assisted in the research and the preparation of the work should be included in an acknowledgment section, which is placed just before the reference section in your document.

>>>>

	\renewcommand{\shortauthors}{Kumar and Eyraud-Dubois, et al.}

	%
	% The abstract is a short summary of the work to be presented in the article.
	\begin{abstract}
		With increasing complexity of hardwares, systems with different memory nodes are ubiquitous in high performance computing (HPC). It is paramount to develop strategies to overlap the data transfers between memory nodes with computations in order to exploit the full potential of these systems. In this paper, we consider the problem of deciding the order of data transfers between two memory nodes for a set of independent tasks with the objective to minimize the makespan. We prove that with limited memory capacity, obtaining the optimal order of data transfers is an NP-complete problem. We propose several heuristics for this problem and provide details about their favorable situations. We present analysis of our heuristics on traces, obtained by running 2 molecular chemistry kernels, namely, Hartree–Fock (HF) and Coupled Cluster Single Double (CCSD) on 10 nodes of an HPC system. Our results show that some of our heuristics achieve significant overlap for moderate memory capacities and are very close to the lower bound of makespan.
	\end{abstract}

@@

	With the advent of multicore, and the use of accelerators, it is notoriously cumbersome to exploit the full capability of a machine. Indeed, there are several challenges that come into picture. First, every architecture provides its own efficacy and interface. Therefore, a steep learning curve is required for programmers to take good utilization of all resources. Second, scheduling is a well known NP-Complete optimization problem, and hybrid and distributed resources make this problem harder (we refer ~\cite{webpagescheduling} for a survey on the complexity of scheduling problems and ~\cite{bleuse2015scheduling} for a recent survey in the case of hybrid nodes). Third, due to shared buses and parallel resources, it is challenging to obtain a precise model based on prediction of computation and communication timings. Fourth, then number of architectures has increased drastically in recent years, therefore it is almost impossible to develop hand tuned optimized code for all these architectures. All these observations led to the development of different task based runtime systems. Among several runtimes, we may cite QUARK~\cite{YarKhan:2011:Quark:Manual} and PaRSEC~\cite{parsec} from from ICL, Univ. of Tennessee Knoxville (USA), StarPU~\cite{starpu} form Inria Bordeaux (France), Legion~\cite{legion12} from Stanford Univ. (USA), StarSs~\cite{ompss} from Barcelona Supercomputing Center (Spain), KAAPI~\cite{kaapi} from Inria Grenoble (France). All these runtime systems allow programmers to express their algorithms at the abstract level in the form of direct acyclic graphs (DAG), where vertices represent computations and edges represent dependencies among them. Sometimes some static information is also provided along with the DAG, such as distance to exit (last) node as a priority or affinity of computation towards resources. The runtime is then responsible for managing scheduling of computations and communications, data transfers among different memories, computation-communication overlap, and load balance.

@@

	Certain applications such as dense linear algebra kernels have regular structure. Therefore, it is possible to associate priorities to computations, based on the task graph structure, and to use them at runtime to make the execution efficient. In irregular applications, programmers do not know the precise structure of the task graphs in advance: tasks are added recursively based on certain sentinel constraints. For such applications, the runtime system sees a set of independent ready tasks and schedules them on different processing units. It is extremely important for runtimes to decide the order of data transfers for these scheduled computations so as to maximize the overlap between computations and communications. This is the main topic of this paper. We prove that the order of communications on two memory nodes with the objective of minimizing the makespan is a NP-Complete problem if the memory of the target node is limited. Our proof is inspired from work by~\cite{Papadimitriou:1980:FSL:322203.322213}, which applies a similar technique for 2-machine flowshop problem with bounded capacity. The main difference between both approaches is that they consider all tasks have the same occupation on the second machine and the memory occupation starts when the processing finishes on the first machine. On the contrary, our approach is designed for tasks appearing in scientific workloads whose memory requirements are highly irregular and we consider that memory is acquired before starting the data transfer on communication resource. We propose different runtime strategies in order to maximize the overlap of computations and communications. We evaluate our strategies on the context of a cluster of homogeneous nodes. However, our approach is generic and easily adaptable to any system which operates on different memory spaces. Here are the important contributions of this paper:

@@

	The problem of scheduling tasks has been highly studied in the literature and many formulations are known to be NP-Complete~\cite{GareyJohnson}. Many static and dynamic strategies have been proposed and analyzed for scheduling tasks on heterogeneous resources~\cite{heft-Topcuoglu,hipc16multiresource,ipdps16starpu}. There is also a number of studies in the direction of task scheduling with the emphasis on improving locality and minimizing the communication cost~\cite{starpu,heft-Topcuoglu}. Stanisic et. al~\cite{luka-dmdar} proposed a heuristic to schedule tasks on a computing resource where most of its data is available. A similar approach has been adopted by Agullo et. al for the scheduling of sparse linear algebra kernels~\cite{agullo_fmm}. Predari et. al proposed heuristics to partition the task graph across a number of processors such that inter-processor communication can be minimized~\cite{predari:tel-01518956}.

@@

	\noindent\textbf{Problem $DT$} : A set of tasks $ST=\{T_1,
	\cdots, T_n\}$ is scheduled on a processing unit $P$ with
	memory unit $M$ of capacity $C$. Input data for tasks of $ST$
	reside on another memory unit $M'$. $COMM_i$ is the time to
	transfer input data from $M'$ to $M$ for task $i$ and $COMP_i$
	is the execution time of task $i$ on $P$. We assume that these
	tasks do not produce any output data. There can be only one
	communication at a time, and $P$ can only process one task at
	a time. A task uses an amount of memory in  $M$ from the
	start of its communication to the end of its computation.

@@

	Given a schedule, \scomm($i$) and \scomp($i$) represent the
	start times of task $i$ on communication and computation
	resources. A schedule is feasible if for every time $t$, the
	amount of memory required by all tasks such that $\scomm(i) \leq t
	\leq \scomp(i) + COMP_i$ is not more than the memory capacity
	$C$. 
	\todo[inline]{SKumar: We do not need this assumption everywhere. CommVolume=CommTime is not true for experimental section. We used this assumption only in NP-hard proof.}
	For simplicity, we assume in all of this paper that the
	amount of memory required by a task is proportional to its
	communication time, and without loss of generality we consider
	that memory size of a task is equal to its communication time.

@@

	Algorithm~\ref{alg:OrderOfExecutionInfinteMemory} presented in Section~\ref{sec:theoreticalProof} achieves an optimal makespan when there is no memory constraint. This optimal value indicates a lower bound on the makespan of the constrained case. We denote this value with $optimal\_ \allowbreak makespan\_  \allowbreak infinite\_  \allowbreak memory$ ($OMIM$). In the present Section, we propose different heuristics for the limited memory case, and we assess their efficiency with respect to this lower bound in Section~\ref{sec:exp}.

@@

	\subsection{Static Ordering}
	In this class of strategies, we compute the order of processing in advance based on criteria such as communication time and compute time. After computing the order, we follow the same order on computation and communication resources and make sure that the memory constraint is respected at each point in the schedule.

@@

	In order to highlight the different behaviors of these static strategies, we propose on Table~\ref{tab:staticOrderExample} an example instance, and on Figure~\ref{fig:staticOrderExample} the corresponding schedules for all these heuristics.

@@

	Dynamic strategies are based on the following principle: when the communication resource is idle, a task is chosen based on a ranking criterion which differ depending on the heuristic, among those which fit in memory. For example, if the selection criteria is to choose a highly compute intensive task, then we compute the ratio of compute time and communication time for all tasks, and we select the task which fit in the currently available memory and whose ratio is maximum (ties are broken arbitrarily). If no task satisfies the selection criteria thee we leave the resource idle at that point and proceed to the next event point. We also ensure that the order on both resources is the same, by ordering tasks on the computing resource accordingly.

%%	\todo[inline]{LED: I added this part ``among those which fit in memory'', is it correct? SKumar: Yes, you are correct.}
%%	\todo[inline]{LED: What happens if no task fit in memory? Skumar: look at the next event point :)}

@@

	In all strategies, the initial order is obtained by Algorithm~\ref{alg:OrderOfExecutionInfinteMemory}. We define the following heuristics based on how we select a task from the set of tasks which fit in memory. If not any task fits in the memory then we leave the resource idle and forward to the next event point.
%%	\todo[inline]{LED: What happens if no task fit in memory?}

@@

		\todo[inline]{SKumar: Move to an extended version and more clear about results and implementation}
		Figure~\ref{fig:iterativeLpSolution} shows the performance of different heuristics with MILP based heuristics for various memory capacities of a single trace file.
		%%	\subsection{MILP Based Strategy}
		%%We use GLPK solver v4.65 to solve the MILP formulation described in Section~\ref{subsec:linearprogrammingformulation}. The solver was unable to handle number of variables required to solve MILP at the scale of our interest. Hence, we solve the linear program iteratively for a small subset of size $k=3,4,5$ (solver was unable to handle number of variables required for $k\ge6$ ) and represent the makespan calculated by this heuristic as $lp.k$. At the boundary of two iterations we fixed the event (communication or computation) of an unfinished task who has started before the boundary point and consider other events flexible. We compute various $lp.k$ values for different memory capacities and observe that most of the other heuristics perform better than this heuristic. Figure~\ref{fig:iterativeLpSolution} shows the comparison of different heuristics with MILP based heuristic for various memory capacities of a single trace file.
		\todo[inline]{SKumar: Figure~\ref{fig:iterativeLpSolution} would be removed and I will redraw this or similar figure for extended version}
		\begin{figure}[htb]
			\includegraphics[scale=0.5]{./results/makespan_with_lp.pdf}
			\caption{Comparision of different heuristics with MILP solution based heuristic.\todo[inline]{LED: I understand that this figure will be removed? Otherwise, use aes(fill=) instead of aes(color=) to have color filled bars.}}
			\label{fig:iterativeLpSolution}
		\end{figure}

@@

		Here moderate memory capacity refers to the case where memory is limited but close to the total memory requirement of all tasks in $OMIM$. We can clearly observe the favorable scenarios for a few heuristics in Figures~\ref{fig:ratio_to_optimal_hf} and ~\ref{fig:ratio_to_optimal_ccsd}. For example, HF compute intensive tasks have small communication timings, hence $SCMR$ heuristic exhibits very good performance in limited memory cases. CCSD has significant percentage of both types of tasks, therefore $OOLCMR$ and $OOSCMR$ heuristics performance is very close to optimal in moderate memory cases. 
		\todo[inline]{Skumar: Initially I placed this table after results. Then I realized that this is our hypothesis not conclusion. Only few favorable scenarios are matched with results. We do not have enough results to show favorable scenarios for all of them.}
		\todo[inline]{LED: Since this is based on experimental results, should this go to a Section 5.5 or 5.6?}

@@

		We consider a machine called Cascade~\cite{Cascade}, available at PNNL, for our experiments. We obtain traces by running two molecular chemistry applications, double precision version of Hartree-Fock (HF) and Coupled Cluster Single Double (CCSD) of NWChem\todo{cite or url} package on 10 nodes of this machine. Each node is composed of 16 Intel Xeon E5-2670 cores. NWChem takes advantages of a Partitioned Global Address Space Programming Model called Global Arrays (GA)~\cite{GlobalArray} to use shared-memory programming APIs on distributed memory computers. GA dedicates one core of each node to handle other cores, hence we can view a node as being composed of 15 computational cores. We use 150 processes for each application and obtain 150 trace files. We run CCSD with Uracil molecules input and HF with SiOSi molecules (for Uracil molecules, HF has a much smaller workload, each processor executes only around 20 tasks, that is why we chose SiOSi input for HF execution). Each process executes around 300-800 tasks. Our data transfer model is quite simple and we consider that all data transfers between the local memory of each process and the GA memory take the same route. Modeling of different routes of data transfers for the same source-destination pair, bandwidth sharing for different source-destination pairs and network congestion is more challenging and part of our future work. This simple model is enough to provide insight to the application developers (or runtime system) about the ordering of data transfers for the same source-destination pair so as to maximize communication-computation overlap. Our model is easily adaptable to any source-destination pair when there is one fixed route between source and destination (such as between CPU and GPU, one copy engine to transfer data from CPU (resp. GPU) to GPU (resp. CPU) ).
		\todo[inline]{More about tiling}
		Both applications mainly perform two types of computations, tensor transpose and tensor contraction. HF expects to specify a tile size and we set it to 100, so that each core can be used efficiently. CCSD automatically determines tile sizes at different program points based on the input molecules. Hence, HF operates on almost homogeneous tiles while CCSD uses more heterogeneous tiles.

@@

		\todo[inline]{SKumar: will do}
		\begin{figure}[htb]
			%\begin{tabular}{c}
			\includegraphics[scale=0.5]{./results/plots/application_properties.pdf}
			\caption{HF and CCSD tasks characteristics.\todo[inline]{ylab(``Ratio to optimal'') (or to OSIM). Ideally remove underscores from the X axis.}}
			\label{fig:ApplicationProperties}
		\end{figure}

@@

		To get more insights into the considered workloads, we provide in Figure~\ref{fig:ApplicationProperties} information about
		the instances we consider. For each instance, we compute the sum of communication times (\textit{sum comm}) and sum of computation times  (\textit{sum comp}), and normalize it relatively to the $OMIM$ value. Figure~\ref{fig:ApplicationProperties} also shows the maximum of both values, which is a lower bound on the possible makespan of a schedule for this instance, and their sum, which is an upper bound: this represents the makespan of the sequential schedule, obtained with zero overlap between computation and communication. We can see that HF is a communication intensive application and at most 20\% overlap can be expected in the best scenario. On the other hand, in the CCSD workload, communications and computations are almost evenly distributed and a more significant overlap is possible.

@@

		%%	\todo[inline]{Scheduled based on Gilmore-Gomory algorithm and its performance }
		We also consider two other static heuristics for evaluation. The first heuristic is based on an algorithm, proposed by Gilmore and Gomory, to obtain the minimal cost sequence for a set of jobs~\cite{Gilmore-Gomory:1964}. This is a classic algorithm for 2-machine no-wait flow shop. In this algorithm, each job has a start and end state and a cost is associated to change the state. In our context, this cost can be seen as non-overlap time of computation for two adjacent tasks. Here is the main idea of this algorithm. Initially, a partial sequence of jobs is represented by a graph such that their overlap is maximum. Subsequently edges are greedily added to this graph to connect two components while minimizing the total non-overlap cost. When the graph is  connected, an edge interchange mechanism is used to determine the sequence of jobs, which ensures that the sequence has minimal cost. More details can be found in the original paper~\cite{Gilmore-Gomory:1964} and our implementation is publicly available~\cite{gitworkrepo}. This algorithm does not take memory constraints into account and only provides the sequence of processing. We use this sequence with a memory capacity restriction just like for other static heuristcs, and we call this heuristic \textit{Gilmore-Gomory} ($GG$).

@@

		\subsection{Experimental results}
		We evaluate our scheduling heuristics for several memory capacities. From the obtained traces, we first determine the minimum requirement of the memory capacity $m_c$ to execute all tasks. Then we observe the behavior of all heuristics with memory capacity $m_c$ to $2m_c$, in increments of $0.125m_c$. Our performance metric is the ratio to optimal $r$: if heuristic $H$ has makespan $M_H$ on an instance, and the optimal makespan for the infinite memory case is $OMIM$, then $r(H)=\frac{M_H}{OMIM}$ (lower values are better). This ratio is at least $1$, and a value close to $1$ indicates a well-suited  heuristic which achives maximum possible communication-computation overlap.

@@

		As indicated above, HF tasks operate on less heterogeneous tiles, this is also noticeable in Figure~\ref{fig:ratio_to_optimal_hf}. All heuristics depict similar behavior for minimum memory capacity $m_c$ and increasing the memory capacity slightly does not change the performance of all heuristics. As memory capacity is increased further, dynamic variants of heuristics start performing better. For the moderate memory capacities (close to $2m_c$), static order with dynamic correction variants outperform others. $GG$ heuristic does not achieve good performance, because its task sequence is obtained considering no extra memory is available, but is then applied in a different scenario where memory is limited. Surprisingly, the $BP$ heuristic which considers only memory constraint obtains good performance for a static heuristic, but is outperformed by more dynamic approaches.
%%		\todo{you mean infinite memory? SKumar: GG is classic in no-wait (or no buffer) 2-macine flowshop}

@@

		Figure~\ref{fig:ratio_to_optimal_best_hf} shows the performance comparison of the best variant in each category, in addition to the \textit{order of submission} ($OS$) strategy which orders tasks in the (arbitrary) order in which they are given. Static strategies are expected to perform better when there is not any memory capacity restriction, and indeed this plot shows that static strategies face capacity bottleneck and underperform with limited memory. Dynamic strategies achieve better performance with limited memory capacity, and static order with dynamic correction strategies are suitable for scenarios with moderate memory capacity.

		\todo[inline]{LED: I do not understand, on Figure~\ref{fig:ratio_to_optimal_best_hf} it seems that \textit{static dynamic best} is always the best option.}
		\todo[inline]{SKumar: If you look at Figure~\ref{fig:ratio_to_optimal_best_hf} carefully,  in the beginning dynamics variants outperform others. This behavior is easily noticeable in Figure~\ref{fig:ratio_to_optimal_hf}. }

@@

		Figure~\ref{fig:ratio_to_optimal_best_ccsd} shows that best variants of dynamic and static order with dynamic correction strategies achieve similar performance at minimum memory capacity $m_c$. But as memory capacity increases, heterogeneity allows static order with dynamic corrections based strategies to take advantage of static knowledge to get maximum overlap and dynamic correction to select another task in case of memory capacity limitation. Static strategies also start performing better at the end, which indicates that this application has potential for significant communication-computation overlap and pure dynamic strategies are unable to take this information into account while making scheduling decisions.

@@

		In this paper, we consider the problem of deciding the order of data transfers between two memory nodes such that overlap of communication and computation is maximized. With Exascale computing, applications face bottlenecks due to communications. Hence, it is extremely important to achieve the maximum overlap of computation and communication in order to exploit the full potential of the system. We show that determining the order of data transfers is an NP complete problem. We propose several data transfer heuristics and evaluate them on two molecular chemistry kernels, HF and CCSD. Our results show that some of our heuristics achieve significant overlap and perform very close to the lower bound. We plan to evaluate our strategies on different applications coming from multiple domains. We also plan to study the behavior of our strategies in the context of overlapping CPU-GPU communications with computations. A runtime system aiming at exposing different heuristics to maximize the communication-computation overlap at the developer level and automatically selecting the best one is currently underway.

@@

			We are grateful to Ajay Panyala for his help in installation of  NWChem on Cascade machine and providing internal details about the package and its inputs. This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of Advanced Scientific Computing Research .......
			%% 			Identification of funding sources and other support, and thanks to individuals and groups that assisted in the research and the preparation of the work should be included in an acknowledgment section, which is placed just before the reference section in your document.
